\clearpage
\phantomsection
\addcontentsline{toc}{chapter}{Appendices}
\chapter*{Appendices}

\section*{Appendix A: Project Timeline}
\addcontentsline{toc}{section}{Appendix A: Project Timeline}
The project timeline is shown below as editable Mermaid code.

\begin{lstlisting}[language={},caption={Project Development Timeline in Mermaid},label={lst:gantt}]
gantt
    dateFormat  YYYY-MM-DD
    title Project Development Timeline
    section Data Collection & Prep
    Data Collection      :done,    a1, 2026-01-01, 10d
    Data Cleaning        :done,    a2, after a1, 5d
    Feature Engineering  :active,  a3, after a2, 2026-01-20
    section Model Building
    Model Selection      :done,    a4, after a3, 5d
    Model Training       :active,  a5, after a4, 10d
    Hyperparam Tuning    :         a6, after a5, 7d
    section Deployment & Testing
    Dashboard Development:         a7, after a6, 15d
    User Acceptance Test :         a8, after a7, 7d
\end{lstlisting}

\section*{Appendix B: Sources Consulted}
\addcontentsline{toc}{section}{Appendix B: Sources Consulted}
\begin{itemize}
    \item User-provided report content and structure notes.
    \item Malhotra et al. on customer churn prediction for FinTech companies using artificial neural networks \cite{malhotra2017}.
    \item Imani et al. comparative analysis of Random Forest and XGBoost under imbalance handling strategies \cite{imani2024}.
    \item Imani and Arabnia systematic review on churn prediction trends and challenges \cite{systematicreview}.
    \item Bhattacharjee et al. work on categorical encoding and standard scaling for churn prediction \cite{bhattacharjee2026}.
    \item Interview Query article for general understanding of public churn datasets \cite{interviewquery2025}.
    \item Amazon SageMaker example pipeline for XGBoost churn prediction workflow \cite{sagemakerxgb}.
\end{itemize}

\section*{Appendix C: Final Submission Checklist}
\addcontentsline{toc}{section}{Appendix C: Final Submission Checklist}
\begin{itemize}
    \item Title page, certificate, acknowledgements, executive summary, and abstract included.
    \item Table of contents, list of figures, and list of tables included.
    \item Chapters 1 through 8 completed in formal report format.
    \item System architecture, DFD, UML, and deployment representations included.
    \item Tables for dataset attributes, algorithm comparison, hyperparameters, metrics, and results included.
    \item References and appendices included.
    \item Placeholder figures inserted for ROC curve, confusion matrix, and feature importance charts.
\end{itemize}
