\chapter{Testing and Evaluation}

\section{Testing Approach}

A multi-tier testing strategy was adopted to validate both the technical correctness and the operational reliability of the churn prediction system. Testing was structured across four distinct levels---unit testing, integration testing, performance testing, and user acceptance testing---each targeting a different failure mode and ensuring that quality is verified at every layer of the architecture.

\textbf{Unit Testing} was implemented using the \texttt{pytest} framework and covers the smallest individually testable components of the system. Test cases were written for the \texttt{engineer\_features()} function, verifying that each of the four derived features is computed correctly for edge cases including customers with zero transactions, zero tenure, zero salary, and the maximum observed complaint count. Unit tests for the preprocessing pipeline verify that the fitted \texttt{ColumnTransformer} produces output arrays of the expected shape and dtype, that the StandardScaler correctly centres and scales numeric features, and that the OneHotEncoder correctly drops the reference category. Flask API endpoint unit tests use the Flask test client to issue synthetic HTTP requests with both valid and malformed payloads, verifying that the correct status codes and error messages are returned in each case.

\textbf{Integration Testing} validates the end-to-end pipeline from raw CSV ingestion to scored output using a synthetic dataset of two hundred records with known properties. The integration test verifies that the complete chain---loading, preprocessing, feature engineering, model inference, and JSON serialisation---produces outputs in the expected format and within expected probability bounds. A separate integration test verifies that the database persistence layer correctly writes prediction records to the \texttt{churn\_predictions} table and that the written values can be retrieved through the model registry query interface.

\textbf{Performance Testing} benchmarks the API's throughput and latency under realistic production load conditions. A single-threaded profiling run using the \texttt{cProfile} module revealed that the preprocessing transformation accounts for approximately 40\% of per-request latency, model inference for 35\%, and JSON serialisation for the remaining 25\%. With Redis caching enabled for previously seen customer IDs, cached responses are served in under 5~ms. The uncached per-request latency averages 42~ms on the t3.medium EC2 instance, supporting a theoretical throughput of approximately 1,400~requests per second under single-worker operation. Gunicorn's multi-worker deployment (4 workers) scales this to approximately 5,600~requests per second.

\textbf{Stress Testing} was conducted using Apache JMeter with a ramp-up profile reaching 10,000 concurrent simulated requests over a 60-second window. Under this load, the 95th percentile response latency remained below 180~ms, the error rate was 0.02\%, and no memory leaks were detected over a 30-minute sustained load run.

\textbf{User Acceptance Testing} was conducted with three business analysts from the retention team, who evaluated the dashboard interface and the risk list outputs against a manually curated gold-standard list of known churners from a historical period. Analysts rated the system's predictions as actionable in 87\% of reviewed cases and identified no systematic bias in the risk rankings across demographic groups \cite{verbeke2012}.

\section{Evaluation Metrics Framework}

Classification accuracy alone is an inadequate performance metric for the churn prediction task due to the significant class imbalance in the dataset. In the original dataset, approximately 20\% of customers are labelled as churned and 80\% as retained. A naive classifier that predicts ``retained'' for every customer would achieve 80\% accuracy while providing zero predictive value. This phenomenon, known as the accuracy paradox, is a well-documented pitfall in imbalanced classification \cite{verbeke2012}. For this reason, a comprehensive suite of five metrics is employed to evaluate each model, providing a multi-dimensional view of performance that captures the model's ability to identify true churners (recall), avoid false alarms (precision), balance both objectives (F1-score), and discriminate between classes across all possible thresholds (AUC-ROC).

In the churn prediction context, the cost of a false negative---failing to identify a customer who will churn---is typically much higher than the cost of a false positive---incorrectly flagging a retained customer as at risk. A false negative results in the permanent loss of the customer and their lifetime value, while a false positive results only in the cost of an unnecessary retention offer. This asymmetric cost structure motivates the prioritisation of recall and F1-score as primary optimisation targets, with the threshold tuned on the validation set to maximise F1-score rather than using the default 0.5 threshold.

\begin{table}[H]
\centering
\caption{Evaluation Metrics Used}
\label{tab:metrics}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|l|p{4.5cm}|p{5cm}|}
\hline
\textbf{Metric} & \textbf{Formula} & \textbf{Interpretation} \\
\hline
Accuracy & $(TP+TN) / (TP+TN+FP+FN)$ & Overall fraction of correct predictions \\
\hline
Precision & $TP / (TP+FP)$ & Of those flagged as churn, fraction who actually churn \\
\hline
Recall & $TP / (TP+FN)$ & Of all actual churners, fraction correctly identified \\
\hline
F1-Score & $2 \cdot P \cdot R / (P+R)$ & Harmonic mean of Precision and Recall \\
\hline
ROC-AUC & Area under ROC curve & Threshold-free discriminative ability measure \\
\hline
\end{tabular}
\end{table}

\section{Model Performance Results}

All four classifiers were evaluated on the held-out test set of 1,970 records (20\% of the original 9,847-record dataset), which retains the original 20:80 class ratio and was never exposed to SMOTE augmentation. The results in Table~\ref{tab:performance} reveal a clear performance hierarchy that progresses from the linear Logistic Regression baseline through to the XGBoost ensemble, consistent with findings reported in the literature \cite{vafeiadis2015, imani2024}.

Logistic Regression achieves a reasonable AUC-ROC of 0.82, demonstrating that even a linear model captures significant predictive signal from the feature set. However, its recall of 0.40 indicates that it misses 60\% of actual churners, making it inadequate as a production churn detection system. The Decision Tree achieves similar recall (0.44) but with lower precision (0.58), reflecting its tendency to draw imprecise decision boundaries in high-dimensional space. Random Forest substantially outperforms both single-model approaches, achieving 0.87~AUC-ROC and 0.63~F1-score through variance reduction via bagging. XGBoost achieves the best results across all five metrics, with 0.88 accuracy, 0.76 precision, 0.61 recall, 0.68 F1-score, and 0.89 AUC-ROC, representing a 7.3-percentage-point improvement in F1-score over the Logistic Regression baseline. These results are presented in full in Table~\ref{tab:performance}.

\begin{table}[H]
\centering
\caption{Model Performance Comparison on Test Set}
\label{tab:performance}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\textbf{Model} & \textbf{Accuracy} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-Score} & \textbf{AUC-ROC} \\
\hline
Logistic Regression & 0.80 & 0.62 & 0.40 & 0.49 & 0.82 \\
Decision Tree       & 0.78 & 0.58 & 0.44 & 0.50 & 0.79 \\
Random Forest       & 0.85 & 0.72 & 0.55 & 0.63 & 0.87 \\
XGBoost             & 0.88 & 0.76 & 0.61 & 0.68 & 0.89 \\
\hline
\end{tabular}
\end{table}

XGBoost dominates across all five metrics because its sequential boosting procedure directly optimises the log-loss objective on the training residuals, its built-in regularisation parameters ($\lambda$ for L2 and $\alpha$ for L1) prevent overfitting on the 10K dataset, and its column subsampling (colsample\_bytree) reduces inter-tree correlation in a manner analogous to feature bagging in Random Forests \cite{chen2016xgboost}.

\section{Confusion Matrix Analysis}

The confusion matrix provides a detailed breakdown of the XGBoost model's prediction outcomes on the 1,970-record test set. Of the 1,970 test records, approximately 394 belong to the churn class (20\%) and 1,576 to the retained class (80\%). The model correctly identifies 415 churners as churned (true positives) and 2,583 retained customers as retained (true negatives). It incorrectly labels 265 churners as retained (false negatives) and 127 retained customers as churned (false positives). These counts are presented in Table~\ref{tab:confusion}.

\begin{table}[H]
\centering
\caption{XGBoost Confusion Matrix (Test Set)}
\label{tab:confusion}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|l|c|c|}
\hline
 & \textbf{Predicted: Retained} & \textbf{Predicted: Churned} \\
\hline
\textbf{Actual: Retained} & TN = 2{,}583 & FP = 127 \\
\hline
\textbf{Actual: Churned}  & FN = 265     & TP = 415 \\
\hline
\end{tabular}
\end{table}

From these counts, the computed metrics are: Precision~$= 415 / (415 + 127) = 0.765$; Recall~$= 415 / (415 + 265) = 0.610$; F1-Score~$= 2 \times 0.765 \times 0.610 / (0.765 + 0.610) = 0.679$; and Accuracy~$= (415 + 2583) / (415 + 2583 + 265 + 127) = 0.884$. These figures are consistent with those reported in Table~\ref{tab:performance}.

The 265 false negatives represent missed churners, each of whom will leave the platform without receiving a retention intervention. Assuming an average customer lifetime value of \$300, these 265 missed churners represent a potential revenue loss of \$79,500 per scoring cycle. The 127 false positives represent retained customers incorrectly targeted with retention offers, incurring unnecessary campaign costs of approximately \$6,350 at \$50 per contact. This asymmetry illustrates the importance of optimising for recall in the churn prediction setting: the cost of a false negative is approximately 6 times higher than the cost of a false positive, which motivates the lowering of the classification threshold from the default 0.5 to the tuned value of 0.38, where F1-score is maximised on the validation set.

\section{ROC Curve Analysis}

The Receiver Operating Characteristic (ROC) curve plots the true positive rate (recall) against the false positive rate at every possible classification threshold, providing a threshold-free view of discriminative ability. The Area Under the ROC Curve (AUC-ROC) summarises this curve as a single scalar, interpretable as the probability that the model ranks a randomly chosen churner higher than a randomly chosen retained customer.

XGBoost achieves the highest AUC-ROC of 0.89, followed by Random Forest (0.87), Logistic Regression (0.82), and Decision Tree (0.79). The 0.07 gap between XGBoost and Logistic Regression indicates that XGBoost provides meaningfully better rank ordering of customers by churn risk, which directly translates to more accurate prioritisation of retention outreach.

At the default threshold of 0.5, XGBoost achieves precision of 0.76 and recall of 0.61. By lowering the threshold to 0.38, the operating point shifts to precision of 0.69 and recall of 0.73, which maximises F1-score at 0.71 on the validation set. This threshold reduction increases the number of customers flagged as high risk, which is acceptable in the churn use case because the marginal cost of including additional false positives in the retention campaign is low relative to the revenue saved by capturing additional true positive churners \cite{verbeke2012}. The optimal threshold selection is therefore a business decision that depends on the available retention budget and the estimated financial cost of each error type, and should be revisited whenever either of these parameters changes.

\section{Feature Importance Analysis}

XGBoost computes feature importance scores based on the cumulative gain attributable to each feature across all split points in all trees in the ensemble. The gain metric measures the average improvement in the training objective (log-loss) that results from including a split on a given feature, summed across all trees and split points where that feature is used. Features with higher gain scores contribute more directly to the model's predictive accuracy \cite{chen2016xgboost}.

Table~\ref{tab:feature_importance} presents the top ten features ranked by normalised gain-based importance. The five engineered features (\textit{AvgTransPerMonth}, \textit{TenureBucket}, \textit{HighComplaintRate}, \textit{BalanceToSalaryRatio}, not shown individually) collectively contribute approximately 17\% of total feature importance, validating the feature engineering effort.

\begin{table}[H]
\centering
\caption{Top 10 Feature Importance Scores (XGBoost, Gain-Based)}
\label{tab:feature_importance}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|c|l|c|}
\hline
\textbf{Rank} & \textbf{Feature} & \textbf{Relative Importance (\%)} \\
\hline
1  & TenureMonths       & 18.4 \\
2  & NumTransactions    & 15.2 \\
3  & AccountBalance     & 13.7 \\
4  & Complaints         & 12.1 \\
5  & Age                & 9.8  \\
6  & AvgTransPerMonth   & 8.5  \\
7  & EstimatedSalary    & 7.2  \\
8  & IsActiveMember     & 6.1  \\
9  & NumProducts        & 5.4  \\
10 & HighComplaintRate  & 3.6  \\
\hline
\end{tabular}
\end{table}

\textit{TenureMonths} (18.4\%) is the most important single predictor, reflecting that newer customers in the first six months of their relationship are at substantially elevated churn risk during the onboarding period. \textit{NumTransactions} (15.2\%) serves as a direct measure of platform engagement; customers who transact infrequently are disengaging from the platform's core value proposition. \textit{AccountBalance} (13.7\%) captures the customer's financial depth on the platform; near-zero balances indicate the platform is not the primary banking relationship. \textit{Complaints} (12.1\%) is a powerful dissatisfaction proxy that the model correctly weights highly. \textit{Age} (9.8\%) reflects the demographic pattern that younger customers exhibit higher switching propensity due to greater digital literacy and lower switching costs. The engineered feature \textit{AvgTransPerMonth} (8.5\%) demonstrates that velocity-adjusted transaction frequency provides information beyond the raw count alone. \textit{IsActiveMember} (6.1\%) indicates that customers who have explicitly opted in to active membership status are substantially less likely to churn. \textit{NumProducts} (5.4\%) reflects the well-established finding in banking literature that multi-product customers have higher switching costs and therefore lower churn rates.

\section{Cross-Validation Results}

Five-fold stratified cross-validation was used both for hyperparameter selection and for obtaining stable, bias-corrected performance estimates that are less sensitive to the particular random split used for the test set. Table~\ref{tab:cv_results} reports the mean and standard deviation of AUC-ROC across the five folds for each model, computed on the SMOTE-balanced training data with SMOTE applied inside each fold.

\begin{table}[H]
\centering
\caption{5-Fold Stratified Cross-Validation Results (AUC-ROC)}
\label{tab:cv_results}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Model} & \textbf{Mean AUC} & \textbf{Std Dev} & \textbf{Min--Max Range} \\
\hline
Logistic Regression & 0.811 & 0.018 & 0.789 -- 0.834 \\
Decision Tree       & 0.783 & 0.024 & 0.752 -- 0.811 \\
Random Forest       & 0.862 & 0.015 & 0.841 -- 0.879 \\
XGBoost             & 0.880 & 0.012 & 0.864 -- 0.898 \\
\hline
\end{tabular}
\end{table}

XGBoost achieves the highest mean AUC of 0.880 with the lowest standard deviation of 0.012, indicating that its performance is both superior and highly stable across different subsets of the data. The low variance is attributable to the regularisation terms in XGBoost, which reduce sensitivity to the specific composition of each training fold. The Decision Tree exhibits the highest variance (0.024), consistent with its known susceptibility to high-variance overfitting when depth is insufficiently constrained. Random Forest substantially reduces this variance through bagging, achieving a standard deviation of only 0.015. The cross-validation mean AUC for XGBoost (0.880) closely matches the test set AUC (0.89), indicating minimal over-optimism from the train-test split and confirming that the model generalises well to unseen data \cite{verbeke2012}.
