\chapter{Conclusion and Future Work}

\section{Summary of Key Findings}

This project has demonstrated that machine learning, and specifically a SMOTE-augmented XGBoost pipeline with domain-informed feature engineering, provides a robust and production-ready solution to the customer churn prediction problem in FinTech platforms. The system was developed and evaluated on a dataset of 9,847 customer records, achieving a test set accuracy of 88\%, AUC-ROC of 0.89, F1-score of 0.68, precision of 0.76, and recall of 0.61 with the tuned XGBoost classifier.

Transaction activity---measured through \textit{NumTransactions} and the engineered \textit{AvgTransPerMonth} velocity feature---emerged as the most important behavioural predictor cluster, collectively accounting for approximately 24\% of total feature importance. The depth of the customer relationship, captured by \textit{TenureMonths}, contributed the highest individual importance score at 18.4\%, reinforcing the finding that relationship duration is the single strongest structural predictor of customer loyalty in digital banking contexts. The complaint count feature contributed 12.1\% of importance, confirming that unresolved service dissatisfaction is a powerful and actionable churn driver.

SMOTE oversampling effectively addressed the 20:80 class imbalance present in the training data, enabling all four models to learn meaningful decision boundaries for the minority churn class rather than defaulting to the trivial majority-class prediction. The comparative evaluation of four algorithms---Logistic Regression, Decision Tree, Random Forest, and XGBoost---provided a comprehensive view of the performance landscape and confirmed the superiority of gradient-boosted ensemble methods for this tabular financial dataset. The four engineered features contributed approximately 17\% of total feature importance, validating the time invested in domain-informed feature construction. The complete pipeline, from data ingestion to scored output through the Flask REST API, is designed for weekly batch deployment and can score a 10,000-customer base in under five minutes on a standard cloud compute instance.

\section{Achieved Objectives}

The following six objectives were defined in Chapter~1. Each has been fully achieved through the work described in this report.

\begin{enumerate}
    \item \textbf{Objective 1 -- Data Acquisition and Preprocessing: Achieved.} A dataset of 9,847 customer records was loaded and preprocessed through a scikit-learn \texttt{ColumnTransformer} pipeline incorporating median imputation for seven numeric features, modal imputation and one-hot encoding for the categorical gender attribute, StandardScaler normalisation, and SMOTE augmentation applied exclusively within the training partition to address the 20:80 class imbalance ratio without introducing data leakage.

    \item \textbf{Objective 2 -- Feature Engineering: Achieved.} Four domain-informed engineered features were derived from the raw attributes: \textit{AvgTransPerMonth} (transaction velocity), \textit{TenureBucket} (customer lifecycle stage encoded as an ordinal variable), \textit{HighComplaintRate} (binary dissatisfaction signal), and \textit{BalanceToSalaryRatio} (financial engagement proxy). These four features collectively accounted for approximately 17\% of total feature importance in the final XGBoost model, confirming their predictive value.

    \item \textbf{Objective 3 -- Model Training and Comparison: Achieved.} Four machine learning classifiers spanning three algorithmic families (linear, tree-based, and boosted ensemble) were trained on the SMOTE-balanced training set and evaluated on the held-out 1,970-record test set. Model performance was quantified using five metrics: accuracy, precision, recall, F1-score, and AUC-ROC. XGBoost achieved the best results across all five metrics.

    \item \textbf{Objective 4 -- Hyperparameter Optimisation: Achieved.} A 5-fold stratified cross-validation grid search was conducted over a predefined parameter space for all four models. For XGBoost, 540 candidate configurations were evaluated. The optimal configuration achieved an AUC-ROC of 0.89 on the test set, representing a 3.5-percentage-point improvement over the default XGBoost configuration evaluated prior to tuning.

    \item \textbf{Objective 5 -- System Deployment: Achieved.} The trained XGBoost model and preprocessing pipeline were serialised using joblib and deployed as a Flask REST API exposing five endpoints for single-record prediction, batch prediction, model metrics retrieval, feature importance retrieval, and high-risk customer listing. A Redis cache layer reduces repeated-query latency to under 5~ms. The deployment architecture on AWS EC2 with RDS PostgreSQL and an Application Load Balancer was specified and documented in Chapter~4.

    \item \textbf{Objective 6 -- Business Impact Quantification: Achieved.} A return on investment analysis using estimated customer lifetime value of \$300 and a retention campaign cost of \$50 per contacted customer demonstrated a net annual benefit of \$46,400 and an ROI of 46.4\% on a targeted campaign using the model's top 20\% risk-ranked customer list. The financial case for deployment is strongly positive under conservative assumptions.
\end{enumerate}

\section{Contributions of this Work}

This project makes five concrete contributions to the practice and research literature on FinTech customer churn prediction.

\textbf{Practical FinTech Churn Prediction Pipeline.} The project delivers a complete, end-to-end machine learning pipeline---from raw CSV ingestion through preprocessing, feature engineering, model training, hyperparameter tuning, and REST API deployment---specifically designed for the FinTech domain. The pipeline is modular, reproducible, and documented at a level of detail sufficient for replication and extension by practitioners.

\textbf{Feature Engineering Methodology.} The project demonstrates the value of domain-informed feature engineering for financial churn prediction. The four engineered features---particularly transaction velocity and tenure lifecycle bucketing---provide predictive signal that is not recoverable from the raw attributes alone, as evidenced by their 17\% combined contribution to feature importance. This methodology is directly transferable to other FinTech sub-domains.

\textbf{Comparative Algorithm Study with Hyperparameter Tuning.} The systematic comparison of four algorithms across a defined hyperparameter search space, evaluated on five metrics, provides FinTech practitioners with a principled framework for algorithm selection in churn prediction tasks. The result that XGBoost consistently outperforms alternatives across all metrics provides actionable guidance for organisations evaluating their modelling approach.

\textbf{Business Impact Quantification.} The explicit translation of model performance metrics (recall, precision) into financial outcomes (retained revenue, campaign cost, net ROI) bridges the gap between data science outputs and business decision-making. This quantification framework can be adapted to any FinTech platform by substituting platform-specific customer lifetime value and campaign cost estimates.

\textbf{Reproducible Experimental Framework.} All experiments are implemented in documented Python code with fixed random seeds, stratified splits, and reproducible preprocessing pipelines, ensuring that results can be independently verified and that the framework can be extended in future work.

\section{Future Work and Enhancements}

The current implementation represents a strong baseline for production deployment, but several high-value enhancements have been identified that would further improve predictive performance, operational robustness, and regulatory compliance.

\begin{enumerate}
    \item \textbf{Real-Time Streaming Scoring.} The current system operates on a weekly batch scoring cadence. Integrating the model with an Apache Kafka streaming pipeline would enable real-time churn scoring triggered by individual transaction events, reducing the latency between a customer's disengagement signal and the retention team's awareness of the risk from days to seconds. This is particularly valuable for high-velocity churn events such as a customer's first complaint or a sudden balance withdrawal, where immediate outreach significantly improves retention rates.

    \item \textbf{Explainable Artificial Intelligence Integration.} Post-hoc explainability is essential for regulatory compliance and for enabling personalised retention conversations. Integrating SHAP (SHapley Additive exPlanations) \cite{lundberg2017shap} would provide per-customer feature attribution scores that explain the specific factors driving each individual's predicted churn probability. LIME (Local Interpretable Model-agnostic Explanations) \cite{ribeiro2016lime} provides a complementary approach based on local linear approximations and would serve as a cross-validation check on the SHAP attributions. Both tools should be integrated into the dashboard to display each high-risk customer's top contributing risk factors alongside their probability score.

    \item \textbf{Adaptive Learning with Drift Detection.} Customer behaviour patterns evolve over time due to macroeconomic changes, platform updates, and competitor actions. A quarterly retraining pipeline with automated Population Stability Index (PSI) monitoring should be implemented to detect distribution drift in both the input feature distributions and the model's output score distribution. When PSI exceeds a predefined threshold (typically 0.25), an automated retraining job should be triggered, the new model evaluated against the current production model on a holdout set, and the better-performing model promoted to production via the model registry.

    \item \textbf{Multi-Source Data Fusion.} The current model relies exclusively on structured transactional and demographic data. Incorporating unstructured data sources---including customer service call transcripts (analysed using natural language processing sentiment classifiers), mobile app session logs (click paths, session duration, feature adoption), and customer satisfaction survey responses---would substantially enrich the feature set. Early experiments in the literature suggest that multi-source models can improve AUC-ROC by 3--5 percentage points over single-source approaches for comparable dataset sizes.

    \item \textbf{A/B Testing Framework.} The current deployment cannot distinguish between customers who were retained because of the intervention and those who would have stayed regardless. Implementing a randomised controlled A/B testing framework---where a random subsample of high-risk customers is assigned to a control group that receives no intervention---would provide a rigorous causal estimate of the retention campaign's true incremental impact, enabling more accurate ROI calculations and preventing overestimation of model value.

    \item \textbf{Advanced Time-Series Feature Engineering.} The current feature set is derived from aggregate statistics over the entire observation window. Computing rolling time-series features---30-day, 60-day, and 90-day transaction trends, balance trajectory slopes, and complaint rate acceleration---would capture the temporal dynamics of disengagement. Graph-based features derived from the customer referral network (degree centrality, clustering coefficient) would capture social influence effects on churn propensity.

    \item \textbf{Model Stacking Ensemble.} A stacking ensemble combining XGBoost, Random Forest, and LightGBM as base learners with a Logistic Regression meta-learner trained on their out-of-fold predictions would likely improve AUC-ROC by 0.5--1.5 percentage points beyond the single best model. Stacking is particularly effective when the constituent models make uncorrelated errors, which is generally the case for algorithmically diverse learners. The LightGBM base learner uses a leaf-wise rather than level-wise tree growth strategy, producing models with different error structures from XGBoost and therefore providing complementary predictive signal.

    \item \textbf{Production Monitoring Dashboard.} A dedicated model operations monitoring dashboard should track data drift (PSI on each input feature distribution), prediction score drift (PSI on the churn score distribution), and model performance degradation (rolling precision and recall on labelled outcomes as they accumulate). Automated alerts should notify the data science team when any monitored metric exceeds a predefined degradation threshold, enabling proactive model maintenance before business impact materialises.

    \item \textbf{Regulatory Compliance and Fairness Auditing.} Financial services regulators in the European Union (under GDPR and the AI Act) and the United States (under ECOA and fair lending regulations) require that automated decision systems do not produce disparate adverse impacts on protected demographic groups. The model should be subjected to a formal disparate impact analysis across the \textit{Age} and \textit{Gender} attributes, computing the ratio of positive prediction rates across demographic groups and flagging any ratio below the 80\% threshold mandated by the four-fifths rule. GDPR-compliant data handling, including the right to explanation for automated decisions and the right to erasure of personal data from the training dataset, must be implemented before the system can be deployed in European markets.
\end{enumerate}

\section{Final Remarks}

The findings of this project carry several concrete recommendations for FinTech organisations considering the adoption of machine learning-based churn prediction systems. First, invest in data infrastructure before model development: the quality and completeness of the feature dataset is the single most important determinant of model performance, and the time invested in robust data pipelines, comprehensive CRM integration, and data governance pays dividends throughout the model lifecycle. Second, prioritise recall over precision in the initial deployment phase, accepting a higher false positive rate to ensure that the system captures the majority of genuine churners; the threshold can be tightened as the retention team scales its operational capacity. Third, plan explicitly for the model lifecycle, including monitoring, retraining, and governance, from the outset rather than treating deployment as the endpoint of the project. Fourth, combine model predictions with human judgement for high-stakes individual retention decisions: the model identifies who is at risk, but the retention conversation is a human interaction that requires empathy, contextual knowledge, and relationship sensitivity.

Customer churn prediction is not a one-time machine learning project but an ongoing MLOps challenge that requires sustained investment in data quality, model monitoring, and organisational capability building. The methods demonstrated in this project---SMOTE-augmented gradient boosting with domain-informed feature engineering, REST API deployment, and business impact quantification---are directly transferable to adjacent FinTech problem domains, including credit default prediction, product cross-sell propensity scoring, fraud detection, and customer lifetime value estimation. The open-source technology stack employed throughout ensures that these methods are accessible to organisations of all sizes and maturity levels, democratising access to production-grade predictive analytics in the financial services sector.
