\chapter{Literature Survey}

\section{Definitions and Importance of Customer Churn}

Customer churn, also referred to as customer attrition or customer defection, describes the process by which a customer terminates their active relationship with a service provider. In the academic literature, a fundamental distinction is drawn between \textit{voluntary churn} and \textit{involuntary churn}. Voluntary churn occurs when the customer makes a deliberate and autonomous decision to discontinue use of the service, typically motivated by dissatisfaction, discovery of a superior alternative, or a change in personal circumstances. Involuntary churn, by contrast, arises from circumstances outside the customer's explicit intent, such as payment failures, account closure by the institution due to policy violations, or the customer's death. The vast majority of churn prediction research, including the present study, focuses on voluntary churn, as this is the category that can be meaningfully influenced by organisational intervention.

The economic significance of churn has been extensively documented across industries. In the telecommunications sector, annual churn rates of twenty to forty percent are commonly reported, representing a persistent and costly drain on operator revenue \cite{verbeke2012}. The financial services industry, while historically characterised by greater customer inertia due to the friction involved in switching banks, has seen churn rates increase substantially with the rise of digital-only providers. Annual churn rates of fifteen to twenty-five percent are now common in traditional banking, while digital banking platforms report rates exceeding thirty percent owing to the ease with which customers can open and close accounts through mobile applications.

The concept of \textit{Customer Lifetime Value} (CLV) provides the most rigorous economic framing for churn. CLV represents the net present value of all future cash flows attributable to a customer relationship, discounted at an appropriate rate. When a customer churns, the organisation does not merely lose the current period's revenue; it forecloses the entire projected stream of future value \cite{reichheld1996}. This makes churn prevention a high-leverage activity: retaining a single high-CLV customer may be worth more than acquiring several new ones, especially when acquisition costs are factored into the equation. The literature consistently demonstrates that churn prediction models, even when imperfect, can generate substantial positive returns when integrated into a disciplined retention programme.

\section{Traditional Machine Learning Models for Churn Prediction}

The application of machine learning to customer churn prediction has a history spanning more than two decades, beginning with the adoption of statistical and linear methods before progressing to more sophisticated non-linear approaches. \textit{Logistic Regression} (LR) remains one of the most widely used baseline classifiers in churn prediction due to its interpretability, computational efficiency, and well-understood probabilistic output. LR models the log-odds of churn as a linear combination of input features, producing calibrated probability estimates that are directly actionable in a business context. Its principal limitation lies in its linearity assumption: it cannot capture complex, non-linear interactions between features without explicit manual engineering of interaction terms.

The \textit{Support Vector Machine} (SVM) was introduced to churn prediction by Guo and Jin \cite{guoen2008}, who demonstrated that the kernel trick enables SVMs to model non-linear decision boundaries in high-dimensional feature spaces by implicitly mapping inputs into a higher-dimensional reproducing kernel Hilbert space. Their work showed that the radial basis function kernel achieved superior separation of churners from non-churners compared to linear classifiers on telecommunications datasets. However, SVMs present challenges in a churn prediction context: they are sensitive to feature scaling, computationally expensive to train on large datasets, and do not natively produce probabilistic outputs, requiring Platt scaling as a post-processing step.

\textit{Decision Trees} (DT) offer a contrasting set of trade-offs. By recursively partitioning the feature space according to information-theoretic criteria such as the Gini impurity or Shannon entropy, decision trees can model arbitrarily complex non-linear boundaries and produce highly interpretable, human-readable rules \cite{malhotra2017}. Their principal weakness is a strong tendency towards overfitting: a fully grown tree memorises the training data and generalises poorly to unseen samples. Pruning strategies and depth constraints mitigate this to some extent, but decision trees on their own remain unreliable predictors in high-dimensional churn datasets with noisy or redundant features.

\section{Ensemble and Boosting Methods}

The recognition of individual classifiers' limitations gave rise to ensemble methods, which combine the predictions of multiple base learners to achieve greater accuracy and robustness than any single model. Breiman's introduction of the \textit{Random Forest} (RF) algorithm \cite{breiman2001rf} represented a landmark advance. Random Forest constructs a large number of decision trees, each trained on a bootstrap sample of the training data and constrained to consider only a random subset of features at each split. The final prediction is obtained by majority vote across all trees. This combination of bagging and random feature selection substantially reduces variance without proportionately increasing bias, and the out-of-bag (OOB) error provides an unbiased estimate of generalisation performance without requiring a separate validation set.

Gradient Boosting Machines (GBM), formalised by Friedman \cite{friedman2001gbm}, take a fundamentally different approach by constructing an ensemble sequentially rather than in parallel. Each successive weak learner is fitted to the negative gradient of the loss function computed on the ensemble's current predictions, allowing the model to progressively correct its errors. XGBoost \cite{chen2016xgboost} extended this framework with several engineering innovations that have made it the dominant tabular data algorithm in competitive machine learning: L1 and L2 regularisation terms added directly to the objective function to control tree complexity, column subsampling analogous to Random Forest's feature bagging, a weighted quantile sketch algorithm for efficient split finding, and native support for parallel tree construction. Empirical benchmarks consistently demonstrate XGBoost's superiority on churn datasets characterised by mixed numerical and categorical features, moderate dataset sizes, and class imbalance \cite{imani2024, bhattacharjee2026}. LightGBM, a more recent gradient boosting implementation, introduces leaf-wise tree growth and histogram-based feature binning to achieve further speed improvements at very large dataset scales, though the performance differential relative to XGBoost diminishes at the ten-thousand record scale typical of this study.

\section{Neural Networks and Deep Learning Approaches}

The application of neural networks to customer churn prediction extends from early work with \textit{Multi-Layer Perceptrons} (MLP) to contemporary deep architectures. Malhotra and Malhotra \cite{malhotra2017} demonstrated that artificial neural networks could outperform logistic regression and decision trees on telecommunications churn datasets, attributing the improvement to the ability of hidden layers to learn hierarchical feature representations that capture non-linear interactions invisible to linear classifiers. Their architecture typically employed two to three hidden layers with sigmoid or ReLU activations, with dropout regularisation applied to mitigate overfitting.

For FinTech churn prediction, the sequential nature of transaction data motivates the use of recurrent architectures. \textit{Long Short-Term Memory} (LSTM) networks model temporal dependencies across transaction sequences by maintaining a cell state that selectively retains or forgets information across time steps through learned gating mechanisms. This enables the model to capture behavioural drift over time, such as a gradual reduction in transaction frequency or a shift towards lower-value transactions, that is difficult to encode in static feature vectors. Attention mechanisms further enhance this capability by allowing the model to weight different time steps differentially based on their relevance to the churn prediction task.

More recently, transformer-based architectures pre-trained on large corpora have been fine-tuned for structured customer data, though their advantages over gradient boosting on purely tabular datasets remain contested \cite{bhattacharjee2026}. The principal limitations of deep learning approaches in the churn prediction context are threefold: they require substantially more labelled training data to generalise reliably, they entail significantly greater computational training costs, and they present a profound interpretability gap that is increasingly relevant in regulated financial services environments where model decisions may be subject to regulatory audit.

\section{Handling Class Imbalance in Churn Datasets}

Class imbalance is a near-universal characteristic of churn datasets. In typical FinTech deployments, churned customers represent between ten and twenty-five percent of the total customer base, meaning that a naive classifier that labels all customers as non-churners can achieve accuracy of seventy-five to ninety percent while providing zero predictive value. This \textit{accuracy paradox} makes standard accuracy an entirely misleading metric in the churn context \cite{burez2009}.

The \textit{Synthetic Minority Over-sampling Technique} (SMOTE), introduced by Chawla and colleagues \cite{chawla2002smote}, is the most widely adopted technique for addressing class imbalance in machine learning. Rather than simply duplicating existing minority class samples, SMOTE generates synthetic examples by interpolating in the feature space between each minority sample and its $k$-nearest minority-class neighbours, where $k$ is a user-specified parameter typically set to five. The resulting synthetic samples populate regions of feature space near the original minority samples, encouraging the classifier to learn a broader and more generalisable decision boundary for the minority class. Variants including ADASYN (Adaptive Synthetic Sampling) and Borderline-SMOTE focus synthetic generation on the most ambiguous regions of the decision boundary, potentially improving performance when misclassification costs are concentrated near the boundary.

Cost-sensitive learning approaches offer an alternative paradigm by modifying the learning objective rather than the data distribution. By assigning higher misclassification penalties to false negatives (missed churners) than to false positives, classifiers such as scikit-learn's implementations of Logistic Regression and Random Forest with \texttt{class\_weight='balanced'} can be induced to prioritise recall on the minority class without requiring oversampling. Ensemble-based approaches such as the Balanced Random Forest further combine resampling with ensemble learning. The critical practical constraint is that any resampling or cost-weighting must be applied exclusively within the training data partition to prevent information leakage into the evaluation set.

\section{Feature Engineering Strategies for Churn Prediction}

Effective feature engineering is arguably the most impactful intervention available to the practitioner building a churn prediction model, often contributing more to predictive performance than the choice of algorithm. The \textit{RFM framework}, originating in direct marketing, encodes customer behaviour along three axes: Recency (how recently the customer has transacted), Frequency (how often they transact), and Monetary value (the scale of their transactions) \cite{imani2024}. These three dimensions capture complementary aspects of customer engagement and have demonstrated robust predictive power across retail, telecommunications, and financial services domains.

Temporal aggregation features extend the RFM concept by computing behavioural statistics over multiple look-back windows simultaneously. For example, computing the ratio of transactions in the most recent three months to the average over the preceding twelve months captures acceleration or deceleration of engagement that neither a single recency measure nor an overall average would reveal. Ratio features such as the average transaction amount divided by account balance, or the number of service complaints per year of tenure, normalise raw counts for differences in customer scale and longevity, making the features more comparable across the customer base.

Interaction terms capture synergistic effects between features that cannot be represented as independent additive contributions. A customer who is both a long-tenure account holder and has recently filed multiple complaints presents a qualitatively different risk profile than a new customer with complaints or a long-tenure customer without them; the interaction of tenure and complaint rate captures this joint effect. Automated feature selection methods including mutual information scoring, LASSO regularisation, and tree-based feature importances provide principled mechanisms for identifying which engineered features carry the most predictive information, enabling practitioners to reduce dimensionality without sacrificing predictive power.

\section{Evaluation Metrics and Best Practices}

The selection of evaluation metrics for churn prediction models requires careful alignment with the business objective and data characteristics. As established in the preceding discussion of class imbalance, classification accuracy is an inappropriate primary metric when classes are unequally represented. \textit{Precision} measures the proportion of customers predicted to churn who actually do, while \textit{Recall} (sensitivity) measures the proportion of actual churners that the model correctly identifies. These two metrics trade off against each other as the classification threshold is varied: lowering the threshold increases recall at the cost of precision, while raising it increases precision at the cost of recall.

The \textit{F1-score}, defined as the harmonic mean of precision and recall, provides a single summary statistic that balances both concerns and is the standard primary metric in class-imbalanced binary classification contexts \cite{vafeiadis2015}. The \textit{Area Under the Receiver Operating Characteristic Curve} (AUC-ROC) is a threshold-independent measure of the model's ability to discriminate between the two classes across all possible decision thresholds, ranging from 0.5 for a random classifier to 1.0 for a perfect one. Its principal limitation in severely imbalanced settings is that the ROC curve can remain visually impressive even when the model performs poorly at the high-precision end of the precision-recall curve; the \textit{Precision-Recall AUC} (PR-AUC) is therefore preferred as a complementary metric for highly imbalanced datasets.

Business-aligned metrics such as the Expected Maximum Profit (EMP) framework translate the model's probabilistic outputs into direct financial terms by incorporating estimates of the cost of contact and the expected revenue recovery from a successful retention intervention \cite{bhattacharjee2026}. The Kolmogorov-Smirnov (KS) statistic, which measures the maximum separation between the cumulative distribution functions of predicted probabilities for churners and non-churners respectively, is widely used in credit risk and churn scoring applications as a threshold-agnostic rank-ordering metric. Cross-validation strategy is equally important: stratified $k$-fold cross-validation preserves the class ratio in each fold, preventing optimistically biased performance estimates that would result from random splits on imbalanced data.

\section{Explainability and Interpretability in Churn Models}

The deployment of machine learning models in financial services contexts raises substantial questions of accountability and interpretability that go beyond predictive performance. Regulatory frameworks such as the European Union's General Data Protection Regulation (GDPR) and the Fair Credit Reporting Act (FCRA) in the United States impose obligations on institutions to explain automated decisions that affect customers, creating a direct regulatory driver for model interpretability.

Lundberg and Lee introduced \textit{SHAP} (SHapley Additive exPlanations) \cite{lundberg2017shap}, which grounds model explanations in cooperative game theory. SHAP values are computed by treating each feature as a player in a cooperative game and assigning each player a contribution equal to their average marginal contribution across all possible feature coalitions, as formalised in the Shapley value from game theory. This approach satisfies desirable axiomatic properties including efficiency (SHAP values sum to the model output), symmetry (equivalent features receive equal values), and null player (features with no effect receive zero SHAP values). TreeSHAP, a computationally efficient implementation for tree-based models, enables exact Shapley value computation in polynomial rather than exponential time, making it practical for deployment-scale explanation. SHAP force plots, summary plots, and interaction value plots provide a multi-level view of feature contributions at both the individual prediction and global model levels.

Ribeiro and colleagues introduced \textit{LIME} (Local Interpretable Model-agnostic Explanations) \cite{ribeiro2016lime}, which approximates the behaviour of any black-box model in the neighbourhood of a specific prediction using a locally faithful linear surrogate. LIME generates an explanation by sampling perturbed versions of the input, obtaining the black-box model's predictions on those perturbations, and fitting a weighted linear regression to the (perturbation, prediction) pairs with weights proportional to the proximity of each perturbation to the original instance. While LIME is model-agnostic and thus applicable to neural networks and SVMs in addition to tree-based models, it has known limitations in stability: explanations for the same instance can vary across repeated calls due to the stochastic sampling procedure, reducing practitioner confidence. Partial Dependence Plots (PDP) and Individual Conditional Expectation (ICE) plots complement SHAP by visualising the marginal effect of a single feature on model predictions averaged or disaggregated across the dataset.

\section{Summary and Research Gap}

The following table summarises key findings from the surveyed literature, providing a comparative overview of datasets, methodologies, and reported performance across representative studies.

\begin{table}[H]
    \centering
    \caption{Summary of Representative Churn Prediction Studies}
    \label{tab:litreview}
    \renewcommand{\arraystretch}{1.3}
    \begin{tabular}{|p{3.2cm}|p{2.5cm}|p{3.5cm}|p{2.2cm}|}
        \hline
        \textbf{Author / Year} & \textbf{Dataset Domain} & \textbf{Models Evaluated} & \textbf{Best AUC} \\
        \hline
        Malhotra \& Malhotra \cite{malhotra2017} & Telecom & ANN, LR, DT & 0.81 \\
        \hline
        Guo \& Jin \cite{guoen2008} & Telecom & SVM, LR & 0.79 \\
        \hline
        Verbeke et al.\ \cite{verbeke2012} & Financial Services & RF, SVM, LR & 0.83 \\
        \hline
        Vafeiadis et al.\ \cite{vafeiadis2015} & Telecom & SVM, RF, NB, LR & 0.85 \\
        \hline
        Imani et al.\ \cite{imani2023} & Banking & XGBoost, RF, DT & 0.88 \\
        \hline
        Imani et al.\ \cite{imani2024} & FinTech & XGBoost, LGBM, RF & 0.91 \\
        \hline
        Bhattacharjee et al.\ \cite{bhattacharjee2026} & Digital Banking & XGBoost, MLP, LSTM & 0.93 \\
        \hline
    \end{tabular}
\end{table}

Analysis of the surveyed literature reveals three substantive research gaps that motivate the present study. First, the majority of high-performing studies draw on telecommunications datasets, which differ structurally from FinTech transaction data in their feature distributions, churn mechanisms, and class imbalance ratios; relatively few studies use FinTech-specific datasets with the rich transactional feature set characteristic of digital banking platforms. Second, the feature engineering reported in most studies is limited to raw demographic and usage statistics; domain-informed derived features capturing transaction velocity, financial health ratios, and customer lifecycle stage are underexplored. Third, while several studies report strong AUC metrics, few translate model outputs into quantified business impact estimates that allow a practitioner to assess the return on investment of model deployment. This study directly addresses all three gaps by applying a carefully engineered feature set to a FinTech-specific dataset, optimising for business-relevant evaluation metrics, and providing a structured business impact analysis in Chapter Six.
