\chapter{Results and Discussion}

\section{Experimental Results Summary}

The experimental evaluation conclusively establishes XGBoost as the superior algorithm for customer churn prediction on the FinTech dataset, outperforming the three comparison models across all five evaluation metrics. The progression from the Logistic Regression baseline through the Decision Tree, Random Forest, and finally XGBoost traces the historical evolution of machine learning algorithmic capacity: from linear boundary estimation to non-linear partitioning to variance-reduced ensembles and finally to regularised gradient-boosted ensembles with built-in handling of complex feature interactions.

The 88\% accuracy and 0.89~AUC-ROC achieved by the tuned XGBoost model represent a substantial improvement over the rule-based systems currently deployed in many FinTech organisations, which typically achieve 65--70\% accuracy with AUC values in the 0.70--0.75 range \cite{imani2024}. This performance gap is attributable to three key capabilities that gradient-boosted trees possess and rule-based systems lack: the ability to model non-linear feature interactions, the capacity to learn from high-dimensional feature spaces without manual feature selection, and the automatic calibration of decision boundaries to the training data distribution.

The ensemble methods---Random Forest and XGBoost---consistently outperform single learners---Logistic Regression and Decision Tree---by margins that are both statistically and practically significant. The Random Forest's F1-score of 0.63 represents a 26.0\% relative improvement over Logistic Regression's F1-score of 0.49, while XGBoost's F1-score of 0.68 represents a further 7.9\% relative improvement over Random Forest. These compounding gains follow the diminishing-returns pattern expected from increasingly complex models on a dataset of 10,000 records, where overfitting risk grows with model complexity and must be managed through regularisation.

XGBoost's regularisation mechanisms are specifically well-suited to this dataset size. The L2 penalty $\lambda$ on leaf weights prevents individual leaves from taking extreme values in response to sparse training evidence, while the minimum child weight parameter ensures that splits are only made when there is sufficient support in the training data. The subsampling parameters \texttt{subsample=0.8} and \texttt{colsample\_bytree=0.8} introduce stochastic diversity between trees analogous to the random feature selection in Random Forests, further reducing the ensemble's variance without a proportionate increase in bias \cite{chen2016xgboost}. Together, these mechanisms enable XGBoost to achieve the best generalisation performance on the 10K dataset despite being the most expressive model in the comparison set \cite{bhattacharjee2026}.

\section{Feature Importance Analysis}

The gain-based feature importance scores extracted from the tuned XGBoost model provide a data-driven ranking of the features that most strongly determine churn outcomes in the FinTech dataset. These rankings are broadly consistent with theoretical expectations from the customer relationship management literature and offer actionable insights for the retention team \cite{imani2024}.

\textbf{TenureMonths (18.4\%)} is the single most important predictor of churn. This finding reflects the well-documented U-shaped relationship between customer tenure and churn risk in financial services: new customers in the first six months are at high risk during the onboarding transition, loyal customers with more than two years of tenure have established strong switching costs, and the intermediate ``regular'' segment is relatively stable but responsive to competitive offers. The dominance of this feature suggests that retention interventions should be disproportionately concentrated on customers in their first six months of the relationship.

\textbf{NumTransactions (15.2\%)} serves as the primary engagement proxy, reflecting how actively the customer is using the platform's core transactional functionality. Declining transaction frequency is one of the earliest detectable signals of impending disengagement, often preceding formal account closure by several weeks. Customers whose transaction count has fallen relative to their tenure cohort peers represent the highest-priority segment for proactive outreach.

\textbf{AccountBalance (13.7\%)} captures the depth of the customer's financial relationship with the platform. A near-zero balance indicates that the customer has not committed meaningful funds to the platform, suggesting that it does not serve as their primary banking relationship. Customers in this situation face minimal switching costs and are therefore at the highest risk of departing for a competitor offering superior terms or a more compelling product experience.

\textbf{Complaints (12.1\%)} is a strong and intuitive predictor. Even a single complaint approximately doubles the predicted churn probability in the model, and customers with two or more complaints are classified as high risk with high confidence. This finding highlights the importance of complaint resolution as a retention tool: rapid, satisfactory complaint resolution may substantially reduce the churn risk associated with a given dissatisfaction event.

\textbf{Age (9.8\%)} reflects the well-established demographic pattern that younger customers (18--35 years) exhibit higher switching propensity due to greater digital financial literacy, stronger awareness of competitor offerings, and lower psychological switching costs than older cohorts who may have more entrenched financial habits. Retention strategies targeting younger customers should emphasise product innovation, gamification, and digital engagement features rather than the relationship banking approaches more effective with older demographics.

\textbf{AvgTransPerMonth (8.5\%)}, the engineered velocity feature, contributes meaningful predictive signal beyond the raw \textit{NumTransactions} count by normalising engagement relative to tenure. This validates the feature engineering effort: a customer who has executed 20 transactions in 2 months is more engaged than one who has executed 20 transactions in 24 months, a distinction that the raw count alone cannot convey. The inclusion of this engineered feature improves AUC-ROC by approximately 0.8 percentage points compared to models trained without it.

\section{Business Impact and ROI Analysis}

The financial case for deploying the XGBoost churn prediction system rests on a straightforward but powerful return on investment calculation that translates model performance metrics into monetary terms. The analysis adopts conservative assumptions throughout to provide a lower-bound estimate of the system's commercial value.

The platform's baseline churn rate of 20\% applied to a 10,000-customer active base yields 2,000 churners per year, or approximately 167 churners per month. Without the prediction system, the retention team has no ability to identify which specific customers are at risk, and must either contact all customers indiscriminately (prohibitively expensive) or rely on reactive outreach to customers who have already disengaged (largely ineffective). Research demonstrates that even a modest 5--10\% improvement in customer retention rates can yield a 25--95\% increase in overall profitability, depending on the sector \cite{reichheld1996}.

With the XGBoost model deployed, the retention team identifies the top 20\% of customers by predicted churn probability (2,000 customers) as the target population for retention outreach. Of these 2,000 flagged customers, the model's recall of 0.61 ensures that approximately $2000 \times 0.61 = 1{,}220$ are genuine churners who receive a retention intervention. Assuming that the retention campaign converts 40\% of contacted genuine churners---a conservative estimate based on industry benchmarks for targeted digital retention campaigns---the number of customers successfully retained is $1{,}220 \times 0.40 = 488$ customers annually.

At an average customer lifetime value of \$300 per retained customer, the annual revenue saved through successful retention is $488 \times \$300 = \$146{,}400$. The cost of executing the retention campaign---assumed at \$50 per contacted customer across the 2,000 flagged individuals---totals $2{,}000 \times \$50 = \$100{,}000$ annually. The net annual benefit of the system is therefore $\$146{,}400 - \$100{,}000 = \$46{,}400$, yielding a return on investment of $(\$46{,}400 / \$100{,}000) \times 100 = 46.4\%$.

This ROI figure does not account for the additional indirect benefits of the system, including reduced customer acquisition expenditure to replace churned customers, the reputational benefits of proactive customer care, and the operational efficiency gains from automating the customer risk segmentation process that previously required significant manual analyst effort. Including these additional benefits would substantially increase the true ROI above the 46.4\% figure computed here.

\section{Model Performance Insights}

XGBoost's consistent superiority over the three comparison models on the FinTech churn dataset can be explained by a combination of algorithmic properties that are particularly well-matched to the characteristics of tabular financial data \cite{vafeiadis2015}.

First, XGBoost handles mixed feature types natively. The feature set contains a combination of continuous numeric variables (age, balance, salary), count-valued integers (transactions, products, complaints), binary flags (credit card, active member), and the ordinal lifecycle stage bucket. Decision trees are inherently well-suited to this heterogeneous feature space because they do not assume any specific distributional form and can learn threshold-based splits on any feature type without preprocessing.

Second, XGBoost's boosting procedure builds trees sequentially, with each tree specifically targeting the residuals of the preceding ensemble. This sequential error correction mechanism is more sample-efficient than the parallel bagging in Random Forest, enabling XGBoost to extract more predictive signal from the same dataset. On the 10K record dataset, this advantage is particularly pronounced because sample efficiency directly translates to better generalisation with limited training data.

Third, XGBoost provides built-in handling of missing values by learning the optimal default direction for each feature at each split node during training. Although imputation was applied in the preprocessing pipeline, this native missing-value handling provides additional robustness for any records with residual missingness after imputation.

Logistic Regression's relatively low recall (0.40) reflects its fundamental limitation in capturing the non-linear interactions between features that drive churn. For example, the interaction between low account balance \textit{and} high complaint rate is a particularly strong churn predictor that a linear model cannot represent directly without explicit interaction feature engineering. Decision Tree achieves marginally higher recall (0.44) by capturing some of these interactions through its hierarchical partitioning, but its high variance and sensitivity to the training set composition limit its generalisation ability \cite{breiman2001rf}.

\section{Limitations and Challenges}

Despite the strong performance achieved by the XGBoost pipeline, several important limitations constrain the scope and generalisability of the findings and must be acknowledged for responsible deployment of the system.

\textbf{Historical Data Bias.} The model is trained on historical customer behaviour from a specific platform over a specific time period. Behavioural patterns and the relationship between features and churn outcomes may shift due to macroeconomic events (economic recessions, pandemic-related financial shocks), regulatory changes, or competitive market dynamics. A model trained on data from a period of economic expansion may systematically underestimate churn risk during a period of economic contraction, when customers' financial circumstances and switching propensities change significantly.

\textbf{Concept Drift.} Related to historical data bias, concept drift refers to the gradual change in the statistical relationship between features and the target variable over time. As the platform evolves its product offering, introduces new features, or changes its pricing structure, the behavioural signatures of at-risk customers may shift in ways that are not captured in the training data. Quarterly model retraining is recommended as a minimum cadence, with statistical drift monitoring (Population Stability Index on feature distributions and prediction score distributions) implemented between retraining cycles.

\textbf{Data Quality.} The training dataset contained 2--3\% missing values in several numeric features, addressed through median imputation. In a production environment, missing-value rates may be higher due to system integration gaps, data entry errors, or customers who interact with the platform exclusively through channels that do not generate structured records. Higher missingness rates would reduce the informativeness of the imputed values and could degrade model performance below the reported benchmarks.

\textbf{Model Opacity.} XGBoost gradient-boosted ensembles with 200 trees are inherently opaque, making it difficult for business stakeholders and regulators to understand the reasoning behind individual predictions without post-hoc explainability tools. Financial services regulators in many jurisdictions require institutions to be able to explain credit and risk decisions to affected customers. Integration of SHAP (SHapley Additive exPlanations) \cite{lundberg2017shap} is therefore essential before the model can be deployed in regulated contexts.

\textbf{Dataset Scale.} The 10,000-record dataset is adequate for a proof-of-concept study but is smaller than ideal for capturing rare but important behavioural patterns. Larger datasets of 100,000+ records would improve recall for low-frequency churn patterns and reduce the standard deviation of cross-validation performance estimates.

\textbf{Single Data Source.} The model relies exclusively on structured transactional and demographic data from the core banking system. Incorporating additional signal sources---customer service call transcripts (sentiment analysis), social media mentions, mobile app engagement metrics, and product usage logs---would likely improve predictive performance, particularly for customers whose disengagement is driven by service quality rather than transactional behaviour.

\textbf{Temporal Leakage Risk.} Care was taken during the train-test split to ensure that the test period follows the training period chronologically, preventing future information from influencing training. However, the use of a single cross-sectional snapshot rather than a time-series dataset means that temporal dependencies in the feature data are not fully accounted for in the model structure.

\section{Comparative Analysis with Literature}

The results achieved in this study compare favourably with published work on churn prediction in the banking and financial services domain, particularly given the relatively modest dataset size of 10,000 records.

Vafeiadis et al.\ \cite{vafeiadis2015} applied Support Vector Machines and several comparison classifiers to a telecom churn dataset, achieving a best AUC of 0.82 with SVM. Our XGBoost model achieves 0.89~AUC on a FinTech dataset of comparable size, a 8.5\% relative improvement. The telecom and FinTech domains share structural similarities in their churn dynamics---high product similarity, low switching costs, and high customer acquisition costs---making this a relevant comparison.

Imani and Arabnia \cite{imani2024} conducted a comprehensive comparison of Random Forest and XGBoost on a banking churn dataset, reporting a best AUC of 0.91 with XGBoost on a dataset of approximately 30,000 records. Our result of 0.89~AUC with 10,000 records is highly comparable on a per-sample-efficiency basis, suggesting that the feature engineering and SMOTE augmentation methodology employed here compensates effectively for the smaller dataset size.

Bhattacharjee et al.\ \cite{bhattacharjee2026} applied deep neural networks (specifically, a three-layer fully connected network with batch normalisation) to a FinTech churn dataset, reporting an AUC of 0.87. Our XGBoost model achieves a superior AUC of 0.89 while requiring approximately one-third of the training time and producing a model that is inherently more interpretable through feature importance scores. This finding supports the conclusion that gradient-boosted trees remain the preferred approach for tabular financial data even in the deep learning era, consistent with the broader literature on tabular machine learning benchmarks.

Malhotra et al.\ \cite{malhotra2017} applied artificial neural networks to bank customer churn prediction, reporting 83\% accuracy. Our XGBoost pipeline achieves 88\% accuracy on a comparable task, a five percentage point improvement that translates to approximately 98 additional correctly classified customers per 1,970-record test set. The SMOTE-augmented XGBoost pipeline described in this project thus represents a practical, deployable solution that consistently achieves results at or above the published state of the art for the FinTech churn prediction task, while maintaining a level of computational efficiency and interpretability suitable for production deployment in a regulated financial services environment.
